---
title: "Mid-Project Group 3"
output: html_notebook
---
```{r}
# Read CSV from a URL

# Import the dataset from Google Drive
url <- "https://drive.google.com/uc?export=download&id=1d8bEgUisDBY-2qC9Csy4qm3d_th_l9PN"
df <- read.csv(url, na.strings = c("NA", "", "NULL"))

# View the first few rows of the dataframe to check the import
head(df)



```
```{r}
#View the summary of the dataset
summary(df)
```
## Data Exploration
Univariate Analysis


HISTOGRAM (NUMERIC)

```{r}
# Histogram of Screen time hours
hist(df$screen_time_hours, 
     main="Histogram of screen time hour", 
     col="cyan", 
     xlab="Screen time hour", 
     border="black")
```

Observation: Most users had screen time between 3 and 8 hours. Few users had very high usage above 10 hours.

```{r}
# Histogram of   bedtime_hour
hist(df$  bedtime_hour, 
     main="Histogram of bedtime hour", 
     col="cyan", 
     xlab="Bedtime Hour", 
     border="black")
```
Observation: Most people sleep 6–8 hours, some less than 5 or more than 9 

```{r}
# Histogram of age 
hist(df$   age , 
     main="Histogram of age ", 
     col="cyan", 
     xlab="Age", 
     border="black")

```

the most common users are around age of 18


```{r}
 # Density Plot
plot (density(df$  screen_time_hours), 
     main="Density Plot of screen_time_hours", 
     col="GREEN", 
     lwd=3)
```
peaks around 5–7 hours

```{r}
# Density Plot
plot(density(df$bedtime_hour), 
     main="Density Plot of bedtime_hour", 
     col="BLUE", 
     lwd=3)
```
Peak around 22:00 

```{r}
# Density Plot
plot(density(df$age), 
     main="Density Plot of age", 
     col="green", 
     lwd=3)
#
```
Most common user are between age of 20 to 25


```{r}
# Boxplot
boxplot(df$age, 
        main="Box Plot of age", 
        col="cyan")
```
Mostly normal, no outliers spotted

```{r}
# Boxplot
boxplot(df$screen_time_hours, 
        main="Box Plot of screen_time_hours", 
        col="cyan")
```
Outliers exist (high screen time)

```{r}
boxplot(df$bedtime_hour, 
        main="Box Plot of bedtime_hour", 
        col="cyan")
```
Mostly normal, no outliers spotted

Exploring Categorical Data
```{r}
 #Count 
table(df$app_type_most_used)
```

```{r}
#Count of Home Country
table(df$stress_level)
```


```{r}
# Bar plot
barplot(table(df$app_type_most_used), 
        main="Count of app_type_most_used", 
        col=c("cyan", "green"))
```
#Educational apps are most used then gaming and utility respectively

```{r}
# Bar plot 
barplot(table(df$stress_level), 
        main="Count of stress_level", 
        col=c("cyan","red", "green"))

```
Stress level moderate is highly distributed 



Multivariate plots

Scatter plot, pair plot and correlation matrix have been displayed to observe different result

```{r}
plot(df$age, df$screen_time_hours,
     main = "Scatter Plot of Age vs Screen Time",
     xlab = "Age",
     ylab = "Screen Time (Hours)",
     col = "blue",
     pch = 16)
```

outliers spotted here


```{r}
plot(df$age, df$bedtime_hour,
     main = "Scatter Plot of Age vs Bedtime Hour",
     xlab = "Age",
     ylab = "Bedtime Hour",
     col = "darkgreen",
     pch = 16)
```


```{r}
plot(df$screen_time_hours, df$bedtime_hour,
     main = "Scatter Plot of Screen Time vs Bedtime Hour",
     xlab = "Screen Time (Hours)",
     ylab = "Bedtime Hour",
     col = "purple",
     pch = 16)

```

More screen time is basically high bed time hour

Box Plot (Numeric vs. Categorical)
```{r}
boxplot(bedtime_hour
 ~ stress_level, data=df,
        main="Box Plot of Bed time hour by stress label",
        col=c("green", "red"))
```

```{r}
boxplot(screen_time_hours ~ stress_level, data=df,
        main="Box Plot of Screen time  hour by stress_level",
        col=c("green", "red"))
```
In this Box plot more outliers were spotted in moderate stress level.


##Check for missing values
```{r}
is.na(df)
```
Count missing values per column
```{r}
colSums(is.na(df))
```
Check total missing values
```{r}
sum(is.na(df))
```
250 missing values observed
Replace NA with the Median of the Column 1,2,3
```{r}
# Replace NA with the Median of each column
df$screen_time_hours[is.na(df$screen_time_hours)] <- median(df$screen_time_hours, na.rm = TRUE)
df$bedtime_hour[is.na(df$bedtime_hour)] <- median(df$bedtime_hour, na.rm = TRUE)
df$age[is.na(df$age)] <- median(df$age, na.rm = TRUE)

# Print the updated dataframe
print(df[c("screen_time_hours", "bedtime_hour", "age")])


```

Replace Categorical column app_type_most_used NA with Mode
```{r}
# Replace missing values in 'app_type_most_used' with the mode
df$app_type_most_used[is.na(df$app_type_most_used)] <- 
  names(sort(table(df$app_type_most_used), decreasing = TRUE))[1]

# Print only the 'app_type_most_used' column
print(df[c("app_type_most_used")])
```

Remove rows of missing values in Target column stress_level

```{r}
# Remove rows where 'stress_level' is NA
df <- subset(df, !is.na(stress_level))


# Print the 'stress_level' column as part of the data frame
print(df[c("stress_level")])


```
```{r}
# Check for missing values in the final dataset (df)
missing_values <- sum(is.na(df))

# Print the result
if(missing_values > 0) {
  cat("There are", missing_values, "missing values in the dataset.\n")
} else {
  cat("No missing values in the dataset.\n")
}

```

```{r}

class(df)


```



```{r}
# Correlation matrix
correlation_matrix <- cor(df[, c("screen_time_hours", "bedtime_hour", "age")], use = "complete.obs")
print(correlation_matrix)

```

```{r}
# Subset the numerical columns and calculate the correlation matrix
correlation_matrix <- cor(df[, c("screen_time_hours", "bedtime_hour", "age")], use = "complete.obs")

# Print the correlation matrix
print(correlation_matrix)

# Plot the correlation matrix using corrplot
library(corrplot)
corrplot(correlation_matrix, method = "circle")

```
Pair plot of numerical columns

```{r}
# Load the GGally package for pair plots
library(GGally)

# Create the pair plot with color by 'stress_level' (target categorical variable)
ggpairs(df[, c("screen_time_hours", "bedtime_hour", "age")], aes(color = df$stress_level))



```
```{r}
# Check for missing values in the final dataset (df)
missing_values <- sum(is.na(df))

# Print the result
if(missing_values > 0) {
  cat("There are", missing_values, "missing values in the dataset.\n")
} else {
  cat("No missing values in the dataset.\n")
}

```





##Data type conversion using label coding

```{r}





# Label encoding for 'app_type_most_used' (Nominal)
df$app_type_most_used_encoded <- as.integer(factor(df$app_type_most_used))

# Label encoding for 'stress_level' (Ordinal)
df$stress_level_encoded <- as.integer(factor(df$stress_level))

# Print the encoded columns to verify
print(df[c("app_type_most_used_encoded", "stress_level_encoded")])




```




##Outlier detection and capping using IQR method


```{r}
# Function to detect outliers using IQR and return bounds
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  
  # Return a list with outlier detection and bounds
  return(list(outliers = x < lower_bound | x > upper_bound,
              lower_bound = lower_bound,
              upper_bound = upper_bound))
}

# Detect outliers for each column and get bounds
screen_time_outliers <- detect_outliers(df$screen_time_hours)
bedtime_outliers <- detect_outliers(df$bedtime_hour)
age_outliers <- detect_outliers(df$age)

# Count the number of outliers for each column
outliers_screen_time_count <- sum(screen_time_outliers$outliers)
outliers_bedtime_count <- sum(bedtime_outliers$outliers)
outliers_age_count <- sum(age_outliers$outliers)

# Print the count of outliers in each column
print(paste("Number of outliers in screen_time_hours: ", outliers_screen_time_count))
print(paste("Number of outliers in bedtime_hour: ", outliers_bedtime_count))
print(paste("Number of outliers in age: ", outliers_age_count))

# Replace outliers with bounds for 'screen_time_hours'
df$screen_time_hours[df$screen_time_hours < screen_time_outliers$lower_bound] <- screen_time_outliers$lower_bound
df$screen_time_hours[df$screen_time_hours > screen_time_outliers$upper_bound] <- screen_time_outliers$upper_bound

# Print outliers for 'screen_time_hours' column
print(df$screen_time_hours[screen_time_outliers$outliers])

```




##Data transformation (scaling)


```{r}
# Scale the numerical columns: screen_time_hours, bedtime_hour, and age
df$screen_time_hours_scaled <- scale(df$screen_time_hours)
df$bedtime_hour_scaled <- scale(df$bedtime_hour)
df$age_scaled <- scale(df$age)

# Print the first few rows of the scaled data
print(df[, c("screen_time_hours_scaled", "bedtime_hour_scaled", "age_scaled")])

```
##New Dataset after preprocessing 
```{r}
# Remove the specified columns from the dataframe
df_new <- df[, !names(df) %in% c("screen_time_hours", "bedtime_hour", "age", "screen_time_hours_capped", 
                                 "app_type_most_used", "stress_level")]

# Print the cleaned dataframe
print(df_new)



```
```{r}
# Check for missing values in the final dataset (df)
missing_values <- sum(is.na(df))

# Print the result
if(missing_values > 0) {
  cat("There are", missing_values, "missing values in the dataset.\n")
} else {
  cat("No missing values in the dataset.\n")
}

```
Recheked again and ensured My dataset has no missing values. All coloumns are properly encoded and removed outliers and scaled.



```

